Factors affecting health sector expenditure in OECD countries

### 2025-05-03 products/manuscript/Manuscript.qmd
#final edits

# Abstract

_We used Ordinary Least Squares method to analyze the factors affecting per capita health expenditure of a country among the OECD nations during 2010 to 2020. We found that GDP growth, control of corruption, and  unemployment rate among the youth in the labor force have a significant and an important role in determining the amount of money spent on health sector by the country. Foreign Direct Investment had a significant but low magnitude impact. Factors such as income share held by the lowest 20-percent population, trade as a percentage of GDP, access to electricity among rural popuation, and age dependency ration had no statistical significant impact on the health expenditure. We also applied Random Forest model to explain the factors affecting per capita health expenditure of a country. The results of both the OLS model and Random forest was tested using cross-validation and train/test technique, and we found that our OLS model shows a consistent result across all test. Although Random Model was found to have a better explanation capability, the inconsistensies raised a question on it's validity. _

@@

## General Background Information

Health expenditure is an important indicator of a country's committment to citizen's well-being and efficiency of it's healthcare systems. Among the OECD (Organisation for Economic Co-operation and Development) countries, the per capita health expenditure vary significantly ranging from 1,000 USD to over 10,000 [@2021].
Understanding the factors that drive these differences is important for policymakers to optimize healthcare delivery and improving healhcare access. While GDP had been one of the primary determinant of health expending [@Newhouse1977], recent studies suggest that institutional, demographic, and socio-economic factors may also play a crucial role in determination of health expenditure among different countries [@Ziegler2017]. The relationship between health expenditure and their determinants because particularly relevant in the context of recent global challenges. The recent COVID-19 pandemic also exposed the vulnerabilities in global healthcare systems with countries showing high pre-pandemic health expenditure generally demonstrating greater resilience [@WHO2022]

@@

We then ran our initial regression with "Health Expenditure Per Capita $USD" as dependent variable, and variables from each of the category as independent variables in separate regressions. The regression results for this preliminary test can be observed under this [ GitHub](https://github.com/xprasanga/PAUDEL-MADA-project) repository under supplementary-materials.

After observing the relation of almost 60 important variables with our variable of interest "Health Expenditure Per Capita $USD", we filtered few variables worth pursuing, and we will discuss them in the sections that will come.

@@

```{r}
#| label: fig-correlation-p2
#| fig-cap: "Correlation based on 2010 data"
#| echo: FALSE
#| out.width: "100%"
knitr::include_graphics(here("results","figures","correlation_plot2.png"))
```

@@

A special concideration should be given to nations that have an increasing GDP, high corruption, and high unemployment rate, as these nations have lower investments in health sectors. These type of countries should therefore be priotized by world bank, World Health Organization, and other institutions who allocate their resources to help countries while also considering the fact that corruption and high unemployment should also be addressed as they show causal effect.

@@

Our analysis covers 2010-2020, excluding the full impact of the COVID-19 pandemic, which dramatically altered health expenditure patterns worldwide. Our OLS approach assumes linear relationships that may not capture complex interactions present in health systems, and although we employ Random Forest to address this, the "black box" nature of machine learning models limits policy interpretability. As with time-series datasets for countries, our datasets had a lot of missing values for a lot of variables across different countries. This is quite common as countries differ in their time and duration of census and surveys and not all countries folllow the sam procedure to collect informations. Also, not all countries would collect the same set of information. We had to drop the missing values due to which we lost a lot of information which coould have been a valuable source of information.

@@

The data for this research is available at the official World Bank- Data Bank website. It is a free database managed by world bank group and can be accessed anytime. We can apply various filters as per our need and export the data as a .CSV or an .xlsx file.

The dataset can be assessed through this [link.](https://databank.worldbank.org/source/world-development-indicators/Type/TABLE/preview/on#)

# Reference

>>>>

<strong>Abstract</strong>

_Understanding the determinants of health expenditures is a critical job in order to optimize the healthcare systems and policymaking. This study utilizes the Ordinary Least Squares method to analyze the factors affecting per capita health expenditure of a country among the OECD nations during 2010 to 2020. Our study identifies that GDP growth, control of corruption, and  unemployment rate among the youth in the labor force have a significant and an important role in determining the amount of money spent on health sector by the country, while Foreign Direct Investment exhibits a significant but low magnitude impact. Factors such as income share held by the lowest 20-percent population, trade as a percentage of GDP, access to electricity among rural popuation, and age dependency ration showed no statistical significant impact on the health expenditure. We also applied Random Forest model to explain the factors affecting per capita health expenditure of a country. The results of both the OLS model and Random forest was tested using cross-validation and train/test technique, and we found that our OLS model shows a consistent result across all tests. Although Random Model was found to have a better explanation capability, the inconsistensies raised a question on it's validity. Our findings highlight the importance of economic and institutional factors in shaping the health expenditure and offer insights for policimakers aiming to enhance healthcare accessibility and efficiency. _

@@

## Health expenditure as an indicator

Health expenditure is an important indicator of a country's committment to citizen's well-being and efficiency of it's healthcare systems. Among the OECD (Organisation for Economic Co-operation and Development) countries, the per capita health expenditure vary significantly ranging from 1,000 USD to over 10,000 USD [@2021].
Understanding the factors that drive these differences is important for policymakers to optimize healthcare delivery and improving healthcare access. While GDP had been one of the primary determinant of health expending [@Newhouse1977], recent studies suggest that institutional, demographic, and socio-economic factors may also play a crucial role in determination of health expenditure among different countries [@Ziegler2017]. The relationship between health expenditure and their determinants because particularly relevant in the context of recent global challenges. The recent COVID-19 pandemic also exposed the vulnerabilities in global healthcare systems with countries showing high pre-pandemic health expenditure generally demonstrating greater resilience [@WHO2022]

@@

We then ran our initial regression with "Health Expenditure Per Capita (USD)" as dependent variable, and variables from each of the category as independent variables in separate regressions. The motivation here was to find the most influential factor from each category. The regression results for this preliminary test can be found in  [ GitHub](https://github.com/xprasanga/PAUDEL-MADA-project). After observing the relation of almost 60 important variables with our variable of interest "Health Expenditure Per Capita $USD", we filtered few variables worth pursuing, and we will discuss them in the sections that will come.

@@

```{r}
#| label: fig-correlation-p2
#| fig-cap: "Additional correlation based on 2010 data"
#| echo: FALSE
#| out.width: "100%"
knitr::include_graphics(here("results","figures","correlation_plot2.png"))
```

@@

A special consideration should be given to nations that have an increasing GDP, high corruption, and high unemployment rate, as these nations have lower investments in health sectors. These type of countries should therefore be priotized by world bank, World Health Organization, and other institutions who allocate their resources to help countries while also considering the fact that corruption and high unemployment should also be addressed as they show causal effect.

@@

Our analysis covers 2010-2020, excluding the full impact of the COVID-19 pandemic, which dramatically altered health expenditure patterns worldwide. Our OLS approach assumes linear relationships that may not capture complex interactions present in health systems, and although we employ Random Forest to address this, the "black box" nature of machine learning models limits policy interpretability. As with time-series datasets for countries, our datasets had a lot of missing values for a lot of variables across different countries. This is quite common as countries differ in their time and duration of census and surveys and not all countries folllow the same procedure to collect informations. Also, not all countries would collect the same set of information. We had to drop the missing values due to which we lost a lot of information which coould have been a valuable source of information.

@@

The data for this research is available at the official World Bank- Data Bank website accessible [here. ](https://databank.worldbank.org/source/world-development-indicators/Type/TABLE/preview/on#) It is a free database managed by world bank group and can be accessed anytime. We can apply various filters as per our need and export the data as a .CSV or an .xlsx file.
